Deep Learning

### 2015-01-04 2014/nature-review/deepreview.tex
fixed some typos, made some suggestions of changes, and added or set a few references

Deep neural networks that learn many layers of feature detectors have dramatically improved the state-of-the-art for
speech recognition, object recognition, object detection, predicting the activity of drug molecules, and many other
tasks. Deep learning discovers intricate structure in large datasets by using the backpropagation algorithm to compute
error-derivatives for the weights that define each feature detector and then repeatedly changed these weights in
proportion to their derivatives averaged over small batches of training examples.  Deep recurrent neural networks, in which
each layer also receives input from its own state at the previous time-step, can now generate good captions for images
and are competitive for machine translation.

@@

So far, most applications of deep learning use feed-forward neural networks, but for tasks that involve sequential
inputs it is often better to use recurrent neural networks. These are given one term from an input sequence at each time
step and their recurrent internal ``hidden'' state depends on the whole history of the input sequence. Recurrent
networks are very good at predicting the next character in a string\cite{Sutskever} or the next word in a
sequence\cite{Mikolov} but they can also be used for more complex tasks.  After reading an English sentence one word at
a time, for example, an English``encoder'' network can be trained so that the final state vector of its hidden neurons
is a good representation of the thought expressed by the sentence.  This thought vector can then be used as the initial
hidden state of a trained French ``decoder'' network which outputs a probability distribution for the first word of the
French translation. If a particuar first word is chosen and provided as input to the decoder network it will then output
a probability distribution for the second word of the translation and so on until a full stop is chosen~\cite{Vinyals, Montreal}.

@@

This review is organized chronlogically starting with the introduction of the backpropagation algortihm in the
1980's. After describing some of its early successes, we explain the actual and perceived limitations that prevented it
from living up to its early promise for so many years and how these limitatons were eventually overcome.

@@

Patterns can be classified by first converting each raw input vector into a vector of feature activations and then
collecting evidence for each possible class by computing a weighted sum of the features. If good features can be
designed by hand, it is relatively easy to learn class-specific weights that optimize
discrimination~\cite{perceptrons}. Widely-used example of such linear classification methods are logistic
 regression and linear support vector machines~\cite{logisticregression,svm}.

However, even with good hand-crafted features, many interesting tasks cannot be solved by linear classifiers. A common
method to get around this problem is to transform the feature vector into a higher dimensional vector through a
non-linear mapping, exploiting the fact that collections of vectors are more likely to be easily spearated when they
are spread around a high-dimensional space~\cite{cover}. One clever way to do so is to use a ``kernel function'' that
measures the similarity between two vectors.  Using the kernel function, any input vector, ${\bf x}_i$ in the training
set can be used to define a feature whose activation for a new input vector, ${\bf x}_{\rm new}$ is just the
similarity between ${\bf x}_i$ and ${\bf x}_{\rm new}$ . Given a kernel function, it is possible to jointly optimize
the choice of training inputs used to define features and the class-specific weights associated with the
features\cite{}. The training inputs used to define features are called ``support vectors'' and Support Vector
Machines\cite{Vapnik} have been very successful, especially for tasks with a relatively small number of training
examples in which features are individually informative. They do not, however, perform well for tasks such as object
recognition in which the input space is high-dimensional and the input-output mapping is so
complex that computing similarities with a simple kernel function is ineffective.

@@

From the earliest days of pattern recognition\cite{selfridge, rosenblatt} researchers aimed to replace hand-engineered
features by multiple layers of learned features.  This was eventually achieved by using the backpropagation algorithm
which was discovered independently by several different groups in the 1970s and 1980s\cite{Werbos, Parker, Lecun,
  Rumelhart, schmidreference}.  Backpropagation is simply an application of chain rule, which yields an efficient way
of computing the gradient of a smooth objective function with respect to all of the weights in a layered network. The
objective function typically measures the discrepancy between the actual output of the network on a given training
case and the target output for that case.  The key insight is that the error derivatives with respect to the {\it
  outputs} of a layer can be computed by working backwards from the error derivatives with respect to the outputs of
the subsequent layer. Once this backwards recursion has been used to compute error derivatives for the outputs of all
layers, it is easy to compute error derivatives for the parameters (or weights) of each layer (see box 1).

@@

Backpropagation provides an efficient way of computing gradients for all of the weights in a layered network on a
single training case, but there are still many ways to make use of these gradients. The most obvious method is to
average the gradients over all of the cases in the training set and then to change each weight in proportion to its
negative gradient. When the training set contains millions of cases and is highly redundant, however, this is
needlessly slow. It is much faster to compute the average gradient on a small ``mini-batch'' of randomly chosen
training cases and to update each weight by this stochastic estimate of its full gradient times a learning rate which
decays as learning proceeds. Stochastic gradient descent can be improved by using the history of the individual weight
gradients to adapt the learning rate for each weight\cite{bottou-bousquet} or to provide additional increments that
depend on earlier gradients\cite{} but these techniques are still the subject of controversy and are beyond the scope
of this review.  However, it is now widely accepted that some form of stochastic gradient descent is the best bet for
large, redundant datasets.

@@

The hidden layers of a multi-layer neural network learn to represent the network's inputs in a way that
makes it easy to predict the target outputs. This is nicely demonstrated by training a multi-layer neural network to
predict the next word in a sequence from a local context of earlier words. Each context word is presented to the network
as a one-of-N vector in which one component has a value of $1$ and the rest are $0$. The output layer has one unit per
word and is forced to produce a probability distribution over words by using the ``softmax'' function:
\begin{equation}
p_j = \frac{exp(z_j)}{sum_i exp(z_i)}
\end{equation}
where $z_j$ is the total input received by output unit $j$. The hidden layers use either the logistic or tanh non-linearity:
\begin{equation}
y_j = \frac{1}{1+ exp(z_j)}, \ \ \ y_j = \frac{}{}
\end{equation}

The connectivity of the network is restricted so that each of the one-of-N input vectors must first be converted to a
``word-vector'' before being used by subsequent layers to give a high probability to the target next word.  The network
learns word-vectors that contain many active components each of which can be interpreted as a separate feature of the
word. This was first demonstrated\cite{RHW} using sequences of three words, such as (Colin, Mother, Victoria), derived
from two family trees. The learned word vector for Colin contained ``semantic'' features representing which family tree
Colin was in, what branch of that tree he was in, and what generation he was\cite{hinton86}. These semantic features
were not explicitly present in the input.  They were discovered by the learning procedure as a good way of factorizing
the structured relationships between the input and output symbols into multiple ``micro-rules''.  For example, given
Colin's generation and the learned feature of Mother that the output should be one generation higher than the input, we
can already restrict the answer to people in a specific generation. The ability of deep learning to discover many such
micro-rules allows it to generalize well to novel combinations of individually familiar context words.

The family tree demonstration used a small, clean dataset in which the regularities could be captured by micro-rules
that were never violated and the features were easy to interpret, but learned word-vectors also work very well when the word
sequences come from a large corpus of real text and the individual micro-rules are unreliable\cite{NPLM} . When trained
to predict the next word in a news story, for example, the learned word vectors for Monday and Tuesday are very similar,
as are the word vectors for Vietnam and Iraq.  Vector representations of words learned from text are now very widely
used in natural language applications\cite{Collobert, Mikolov, Socher?}.

The issue of representation lies at the heart of the debate between the logic-inspired and the neurally-inspired
paradigms for cognition. In the logic-inspired paradigm, an instance of a symbol is something whose only property is
that it is either identical or non-identical to other symbol instances. It has no internal structure that is
relevant to its use. By contrast, neural networks use large activity vectors instead of symbols and parallel matrix
operations followed by non-linearities instead of formal rules of inference. With sufficient data, computational power
and depth, the combination of backpropagation for computing gradients and stochastic gradient descent for updating the
weights on connections has proved to be very good at learning appropriate distributed representations. These
representations allow a lot of parallel computation at the feature level so that complex tasks can be solved in a very
small number of sequential steps.

@@

Recurrent neural networks are very powerful dynamical systems, but training them with BPTT proved to be problematic,
because the backpropagated gradients either grow or shrink at each time step, so over many time-steps they typically explode
or vanish\cite{Bengio, schmdhuber}.

@@

The obvious weakness of gradient descent is that it can get trapped in poor local minima, but in high-dimensional
weight-spaces this is much less common than getting stuck close to saddle points where the error surface curves upwards
in most directions but downwards in a few. Recent empirical\cite{Bengio} and theoretical results\cite{Yann} strongly
suggest that in deep networks almost all of these saddle points give very similar performance so it does not matter
which one we find.  This goes a long way towards explaining why stochastic gradient descent works so well for deep
neural networks.

Interest in deep feed-forward networks was revived in 2005\cite{IJCAI} by the introduction of an unsupervised learning
procedure that could create layers of feature detectors without requiring labeled data. The objective in learning each
layer of feature detectors was to be able to reconstruct or model the activities of feature detectors (or raw inputs) in the
layer below.  By ``pre-training''  several layers of progressively more complex feature detectors using this reconstruction
objective, the weights of a deep network could be initialized to sensible values.  A final layer of output units could
then be added to the top of the network and the whole deep system could be fine-tuned using standard
backpropagation\cite{Science}. This worked remarkably well for recognizing handwritten digits especially when the amount
of labeled data was very limited\cite{BengioNips06}.

The first major application of this pre-training approach was in speech recognition and it was made possible by the
advent of fast Graphics Processing Units (GPUs) that were convenient to program.  In 2009, the approach was used to map
short temporal windows of coefficients extracted from a soundwave to a set of probabilities for the various fragments of
speech that might be represented by the frame in the center of the window.  It achieved record-breaking results on a
standard speech recognition benchmark that used a small vocabulary\cite{TIMITpaper} and was quickly developed to give
record-breaking results on a large vocabulary task\cite{Dahl}.  By 2012, versions of the deep net from 2009 were being
developed by many of the major speech groups\cite{SPS} and were already being deployed in Android phones.  For smaller
datasets, unsupervised pre-training leads to significantly better generalization, but once deep
learning had been rehabilitated, it turned out that large labeled datasets did not need the pre-training stage, provided the
scales of the initial random weights were carefully chosen to avoid very large or very small initial gradients.

Interest in recurrent networks (which are deep in time) was revived by ``Long Short Term Memory'' networks (LSTMs) that
use special hidden units whose natural behaviour is to remember inputs for a long time\cite{Hochreiter}.  A
linear ``memory cell'' interacts with other units via gated connections whose weights are multiplied by the state of a
logistic gating unit that has an activation between $0$ and $1$. The input vector has little effect on the state of a
memory cell if the cell's input gate has a low activation so memory cells can learn to selectively ignore the input.  Similarly,
a memory cell has little effect on the output if its output gate has low activation. Each memory cell also has a
connection to itself at the next time-step that has a weight of 1 so it tends to copy its own real-valued state, but
this connection is gated by a ``keep gate'' so the stored value is quickly eroded unless the keep-gate has an activation
near $1$.  The gates themselves have their activities determined by weighted inputs that they receive from the input
vectors and all of the memory cells.

When LSTMs were first described in English\cite{HochSchmid} their complexity was off-putting and, despite their
impressive ability to learn very long-range temporal structure on toy problems, they were largely ignored until they
achieved excellent results in a competition to read cursive hand-writing\cite{Graves}. They have subsequently proved to
be a very effective for tasks that have sequential input.  They are now better than deep feed-forward networks for
speech recognition\cite{gravestimit} and unlike feed-forward nets they can implement an entire speech recognition system
that goes all the way from coefficents derived from the soundwave to the sequence of characters in the transcription~\cite{graves}.
They are also what is currently used for the encoder and decoder networks that perform so well at machine translation.

It is not yet clear which aspects of the complicated architecture of LSTMs are really necessary. Memory cells are a good
way of allowing the backpropagated error derivatives to neither explode nor vanish, but they are not the only way.  For
example, recurrent nets with rectified linear hidden units that use the non-linear function $y=max(0,x)$ are much
simpler than LSTMs and if the recurrent weight matrix is initialized to be close to the identity matrix they work about
as well as LSTMs both for toy problems involving very long-range temporal structure and for real tasks like predicting
the next word in text\cite{LeHinton}.

@@

When deep convolutional networks were applied to a dataset of about a million images from the web that contained a
thousand different classes they achieved spectacular results, almost halving the error rates of the best computer vision
sytems~\cite{kriz}.  Some of the images and the network's predictions are shown in figure ??.  The major factors in this
success were the large labeled dataset~\cite{FeiFe}, the use of fast GPUs, and the convolutional early layers. were.
Rectified linear units made the training much faster and the fully connected final layers were prevented from
overfitting the peculiarities of the training set by randomly omitting hidden units during
training\cite{dropoutJMLR}. Deep convolutional nets have quickly replaced the carefully engineered pipelines full of
hand-designed features developed by the computer vision community and they are now deployed in many computer vision
applications, including image tagging, object detection, face recognition, facial expression recognition and scene parsing.

@@

Deep learning is now the method of choice for many perceptual tasks and is making major inroads in natural language
processing where a huge amount of messy commonsense knowledge is required. Rapid progress is also being made in applying deep

learning, in particular BPTT, to relatively pure symbol processing tasks.  An impressive recent example is the ``Neural
Turing Machine'' which is a recurrent neural network that can choose when and what to write to or read from an external
memory~\cite{NTM}. Among other things, this can learn to ouput a sorted list of symbols when its input consists of an
unsorted sequence in which each symbol is accompanied by a real value that indicates its priority in the list.  To
achieve this it must learn a sorting program purely from examples of the input and output sequences.  A different system
that also uses BPTT can learn to map the sequence of characters in a simple program written in Python to the seqeunce of
characters that the program would print~\cite{Woj}. Rather surprisingly, BPTT can learn about {\it for loops} and can
also learn to perform addition and multiplication of strings contining several decimal digits. Achieving this kind of
behaviour seemed ludicrously optimistic when BPTT was first proposed.

@@

Human vision is an active process in which intelligent fixations are used to ensure that only a tiny fraction of the optic array is ever processed at high resolution. We expect much of the future progress in vision, especially for video understanding or robotics, to come from systemsthat are trained end-to-end and combine deep convolutional networks with recurrent nets that use reinforcement learning to decide where to look\cite{}.  Such systems already outperform passive vision systems\cite{ba\&mnih}.  Similarly, when understanding sentences or whole documents we expect future systems to learn strategies for selectively attending to one part at a time\cite{Yoshua translation}. We also expect that in the next few years deep recurrent neural networks will become far better at understanding the meaning of documents, including the way in which the thoughts expressed by the sentences are related to one another.

This review has focused on supervised learning because of its recent achievments and we have ignored much
interesting work on unsupervised learning procedures for deep neural networks~\cite{deep-boltz, wake-sleep, google-cat, denoising-auto, ranzato-pami, more recent}. In the long run, however, we believe unsupervised learning will be essential for allowing deep networks to generalize well by creating internal representations that untangle the many different factors that interact to produce images, speech  and documents~\cite{Yann, Yoshua: untangling papers}.

Finally, we believe that the recent successes of deep learning using stochastic gradient descent are only the beginning. 
End-to-end learning systems that have many parameters and few built-in assumptions can make use of more data and more compute power far more easily than systems that require hand-engineering of domain-specific knowledge.

>>>>

Deep neural networks that learn many layers of feature detectors have
dramatically improved the state-of-the-art for speech recognition, object
recognition, object detection, predicting the activity of drug molecules,
and many other tasks. Deep learning discovers intricate structure in large
datasets by using the backpropagation algorithm to compute
error-derivatives for the weights that define each feature detector and
then repeatedly change these weights in proportion to their derivatives
averaged over small batches of training examples.  Deep recurrent neural
networks, in which each layer also receives input from its own state at the
previous time-step, can now generate good captions for images and are
competitive for machine translation.

@@

So far, most applications of deep learning use feed-forward neural
networks, but for tasks that involve sequential inputs it is often better
to use recurrent neural networks. These are given one element from an input % YB term-->element
sequence at each time step and their recurrent internal ``hidden'' state
depends on the whole history of the input sequence. Recurrent networks are
very good at predicting the next character in a string\cite{Sutskever} or
the next word in a sequence\cite{Mikolov} but they can also be used for
more complex tasks.  After reading an English sentence one word at a time,
for example, an English``encoder'' network can be trained so that the final
state vector of its hidden neurons is a good representation of the thought
expressed by the sentence.  This thought vector can then be used as the
initial hidden state of a trained French ``decoder'' network which outputs
a probability distribution for the first word of the French translation. If
a particular first word is chosen (e.g., according to that probability)
and provided as input to the decoder
network it will then output a probability distribution for the second word
of the translation and so on until a full stop is chosen~\cite{Vinyals,
  Montreal}. 
%YB added a clarification
Overall, this process generates sequences of French words according to
a probability distribution that depends on the English sentence.

@@

This review is organized chronlogically starting with the introduction of
the backpropagation algorithm in the 1980's. After describing some of its
early successes, we explain the actual and perceived limitations that
prevented it from living up to its early promise for so many years and how
these limitations were eventually overcome.

@@

Patterns can be classified by first converting each raw input vector into a
vector of feature activations and then collecting evidence for each
possible class by computing a weighted sum of the features. If good
features can be designed by hand, it is relatively easy to learn
class-specific weights that optimize
discrimination~\cite{perceptrons}. Widely-used examples of such linear
classification methods are logistic regression and linear support vector
machines~\cite{logisticregression,svm}.

However, even with good hand-crafted features, many interesting tasks
cannot be solved by linear classifiers. A common method to get around this
problem is to transform the feature vector into a higher dimensional vector
through a non-linear mapping, exploiting the fact that collections of
vectors are more likely to be easily spearated when they are spread around
a high-dimensional space~\cite{cover}. One clever way to do so is to use a
``kernel function'' that measures the similarity between two vectors.
Using the kernel function, any input vector, ${\bf x}_i$ in the training
set can be used to define a feature whose activation for a new input
vector, ${\bf x}_{\rm new}$ is just the similarity between ${\bf x}_i$ and
${\bf x}_{\rm new}$. Given a kernel function, it is possible to jointly
optimize the choice of training inputs used to define features and the
class-specific weights associated with the features~\cite{svm}. The training
inputs used to define features are called ``support vectors'' and Support
Vector Machines\cite{Vapnik} have been very successful, especially for
tasks with a relatively small number of training examples in which features
are individually informative. They do not, however, perform well for tasks
such as object recognition in which the input space is high-dimensional and
the input-output mapping is so complex that computing similarities with a
simple kernel function is ineffective.

@@

From the earliest days of pattern recognition~\cite{selfridge, rosenblatt}
researchers aimed to replace hand-engineered features by multiple layers of
learned features.  This was eventually achieved by using the
backpropagation algorithm which was discovered independently by several
different groups in the 1970s and 1980s\cite{Werbos, Parker, Lecun,
  Rumelhart, schmidreference}.  Backpropagation is simply an application of
the chain rule, which yields an efficient way of computing the gradient of a
smooth objective function with respect to all of the weights in a layered
network. The objective function typically measures the discrepancy between
the actual output of the network on a given training case and the target
output for that case.  The key insight is that the error derivatives with
respect to the {\it outputs} of a layer can be computed by working
backwards from the error derivatives with respect to the outputs of the
subsequent layer. Once this backwards recursion has been used to compute
error derivatives for the outputs of all layers, it is easy to compute
error derivatives for the parameters (or weights) of each layer (see box
1).

@@

Backpropagation provides an efficient way of computing gradients for all of
the weights in a layered network on a single training case, but there are
still many ways to make use of these gradients. The most obvious method is
to average the gradients over all of the cases in the training set and then
to change each weight in proportion to its negative gradient, which is the
direction where a small change will most improve the objective function. 
When the training set contains millions of cases and is highly redundant, however,
this is needlessly slow. It is much faster to compute the average gradient
on a small ``mini-batch'' of randomly chosen training cases and to update
each weight by this stochastic estimate of its full gradient times a
learning rate which decays as learning proceeds. Stochastic gradient
descent can be improved by using the history of the individual weight
gradients to adapt the learning rate for each weight\cite{bottou-bousquet}
or to provide additional increments that depend on earlier gradients\cite{momentum}
but these techniques are still the subject of controversy and are beyond
the scope of this review.  However, it is now widely accepted that some
form of stochastic gradient descent is the best bet for large, redundant
datasets.

@@

The hidden layers of a multi-layer neural network learn to represent the
network's inputs in a way that makes it easy to predict the target
outputs. This is nicely demonstrated by training a multi-layer neural
network to predict the next word in a sequence from a local context of
earlier words~\citep{BenDucVin01-short}. 
Each context word is presented to the network as a one-of-N
vector in which one component has a value of $1$ and the rest are $0$. The
output layer has one unit per word and is forced to produce a probability
distribution over words by using the ``softmax'' function:
\begin{equation}
p_j = \frac{exp(z_j)}{sum_i exp(z_i)}
\end{equation}
where $z_j$ is the total input received by output unit $j$. The hidden
layers use either the logistic or tanh non-linearity:
\begin{equation}
y_j = \frac{1}{1+ exp(z_j)}, \ \ \ y_j = \frac{}{} % YB: what was the intent here for the 2nd part?
\end{equation}

The connectivity of the network is restricted so that each of the one-of-N
input vectors must first be converted to a ``word-vector'' before being
used by subsequent layers to give a high probability to the target next
word.  The network learns word-vectors that contain many active components
each of which can be interpreted as a separate feature of the word. This
was first demonstrated\cite{RHW} using sequences of three words, such as
(Colin, Mother, Victoria), derived from two family trees. The learned word
vector for Colin contained ``semantic'' features representing which family
tree Colin was in, what branch of that tree he was in, and from what generation
he was~\cite{hinton86}. These semantic features were not explicitly present
in the input.  They were discovered by the learning procedure as a good way
of factorizing the structured relationships between the input and output
symbols into multiple ``micro-rules''.  For example, given Colin's
generation and the learned feature of Mother that the output should be one
generation higher than the input, we can already restrict the answer to
people in a specific generation. The ability of deep learning to discover
many such micro-rules allows it to generalize well to novel combinations of
individually familiar context words.

The family tree demonstration used a small, clean dataset in which the
regularities could be captured by micro-rules that were never violated and
the features were easy to interpret, but learned word-vectors also work
very well when the word sequences come from a large corpus of real text and
the individual micro-rules are unreliable\cite{BenDucVin01-short}. When trained to
predict the next word in a news story, for example, the learned word
vectors for Monday and Tuesday are very similar, as are the word vectors
for Vietnam and Iraq.  Vector representations of words learned from text
are now very widely used in natural language applications\cite{Collobert,
  Mikolov, Socher?}.

The issue of representation lies at the heart of the debate between the
logic-inspired and the neurally-inspired paradigms for cognition. In the
logic-inspired paradigm, an instance of a symbol is something whose only
property is that it is either identical or non-identical to other symbol
instances. It has no internal structure that is relevant to its use. By
contrast, neural networks use large activity vectors instead of symbols and
parallel matrix operations followed by non-linearities instead of formal
rules of inference. With sufficient data, computational power and depth,
the combination of backpropagation for computing gradients and stochastic
gradient descent for updating the weights on connections has proved to be
very good at learning appropriate distributed representations. These
representations allow a lot of parallel computation at the feature level so
that complex tasks can be solved in a very small number of sequential
steps. In fact, it can be shown that a measure of the richness of the functions
that can be represented by a deep neural network grows exponentially
with depth~\citep{Montufar-et-al-NIPS2014}.

@@

Recurrent neural networks are very powerful dynamical systems, but training
them with BPTT proved to be problematic, because the backpropagated
gradients either grow or shrink at each time step, so over many time-steps
they typically explode or vanish~\cite{Bengio-trnn93-small,Hochreiter91-small}.

@@

The obvious weakness of gradient descent is that it can get trapped in poor
local minima, but in high-dimensional weight-spaces this is much less
common than getting stuck close to saddle points where the error surface
curves upwards in most directions but downwards in a few. Recent
results~\cite{Dauphin-et-al-NIPS2014-small,Yann} strongly suggest
that in deep networks almost all of these local minima give very similar
performance so it does not matter which one we find.  This goes a long way
towards explaining why stochastic gradient descent works so well for deep
neural networks.

Interest in deep feed-forward networks was revived in 2005\cite{IJCAI} by
the introduction of an unsupervised learning procedure that could create
layers of feature detectors without requiring labeled data. The objective
in learning each layer of feature detectors was to be able to reconstruct
or model the activities of feature detectors (or raw inputs) in the layer
below.  By ``pre-training'' several layers of progressively more complex
feature detectors using this reconstruction objective, the weights of a
deep network could be initialized to sensible values.  A final layer of
output units could then be added to the top of the network and the whole
deep system could be fine-tuned using standard
backpropagation\cite{Science}. This worked remarkably well for recognizing
handwritten digits especially when the amount of labeled data was very
limited\cite{Bengio-nips-2006-small}.

The first major application of this pre-training approach was in speech
recognition and it was made possible by the advent of fast Graphics
Processing Units (GPUs) that were convenient to program~\citep{RainaICML09-small}.  
In 2009, the approach was used to map short temporal windows of coefficients extracted
from a soundwave to a set of probabilities for the various fragments of
speech that might be represented by the frame in the center of the window.
It achieved record-breaking results on a standard speech recognition
benchmark that used a small vocabulary\cite{TIMITpaper} and was quickly
developed to give record-breaking results on a large vocabulary
task~\cite{Dahl}.  By 2012, versions of the deep net from 2009 were being
developed by many of the major speech groups\cite{SPS} and were already
being deployed in Android phones.  For smaller datasets, unsupervised
pre-training leads to significantly better generalization~\citep{Erhan+al-2010-small}, 
but once deep
learning had been rehabilitated, it turned out that large labeled datasets
did not need the pre-training stage, provided the scales of the initial
random weights were carefully chosen to avoid very large or very small
initial gradients.

Interest in recurrent networks (which are deep in time) was revived by
``Long Short Term Memory'' networks (LSTMs) that use special hidden units
whose natural behaviour is to remember inputs for a long
time~\cite{Hochreiter+Schmidhuber-1997}.  
A linear ``memory cell'' interacts with other units
via gated connections whose weights are multiplied by the state of a
logistic gating unit that has an activation between $0$ and $1$. The input
vector has little effect on the state of a memory cell if the cell's input
gate has a low activation so memory cells can learn to selectively ignore
the input.  Similarly, a memory cell has little effect on the output if its
output gate has low activation. Each memory cell also has a connection to
itself at the next time-step that has a weight of 1 so it tends to copy its
own real-valued state, but this connection is gated by a ``keep gate'' so
the stored value is quickly eroded unless the keep-gate has an activation
near $1$.  The gates themselves have their activities determined by
weighted inputs that they receive from the input vectors and all of the
memory cells.

When LSTMs were first described in
English~\cite{Hochreiter+Schmidhuber-1997} their complexity was off-putting
and, despite their impressive ability to learn very long-range temporal
structure on toy problems, they were largely ignored until they achieved
excellent results in a competition to read cursive
hand-writing~\cite{Graves}. They have subsequently proved to be very
effective for tasks that have sequential input.  They are now better than
deep feed-forward networks for speech recognition~\cite{gravestimit} and
unlike feed-forward nets they can implement an entire speech recognition
system that goes all the way from coefficents derived from the soundwave to
the sequence of characters in the transcription~\cite{gravestimit}.  They
are also what is currently used for the encoder and decoder networks that
perform so well at machine
translation~\citep{Bahdanau-et-al-arxiv2014,Sutskever-et-al-NIPS2014}.

% YB: I would skip the next paragraph; we've tried this and failed
% so maybe it is better to wait and see before we put this in a review
It is not yet clear which aspects of the complicated architecture of LSTMs
are really necessary. Memory cells are a good way of allowing the
backpropagated error derivatives to neither explode nor vanish, but they
are not the only way.  For example, recurrent nets with rectified linear
hidden units that use the non-linear function $y=max(0,x)$ are much simpler
than LSTMs and if the recurrent weight matrix is initialized to be close to
the identity matrix they work about as well as LSTMs both for toy problems
involving very long-range temporal structure and for real tasks like
predicting the next word in text\cite{LeHinton}.

@@

When deep convolutional networks were applied to a dataset of about a
million images from the web that contained a thousand different classes
they achieved spectacular results, almost halving the error rates of the
best computer vision sytems~\cite{kriz}.  Some of the images and the
network's predictions are shown in figure ??.  The major factors in this
success were the large labeled dataset~\cite{FeiFe}, the use of fast GPUs,
and the convolutional early layers.  It had already been shown that deep
supervised nets with rectified linear units could be trained without
unsupervised pre-training~\citep{Glorot+al-AI-2011-small}. 
These rectified linear units made the
training much faster and the fully connected final layers were prevented
from overfitting the peculiarities of the training set by randomly omitting
hidden units during training~\cite{dropoutJMLR}. Deep convolutional nets
have quickly replaced the carefully engineered pipelines full of
hand-designed features developed by the computer vision community and they
are now deployed in many computer vision applications, including image
tagging, object detection, face recognition, facial expression recognition
and scene parsing.

@@

Deep learning is now the method of choice for many perceptual tasks and is
making major inroads in natural language processing where a huge amount of
messy commonsense knowledge is required. Rapid progress is also being made
in applying deep learning, in particular BPTT, to relatively pure symbol processing tasks.
An impressive recent example is the ``Neural Turing Machine'' which is a
recurrent neural network that can choose when and what to write to or read
from an external memory~\cite{NTM}. Among other things, this can learn to
ouput a sorted list of symbols when its input consists of an unsorted
sequence in which each symbol is accompanied by a real value that indicates
its priority in the list.  To achieve this it must learn a sorting program
purely from examples of the input and output sequences.  A different system
that also uses BPTT can learn to map the sequence of characters in a simple
program written in Python to the sequence of characters that the program
would print~\cite{Woj}. Rather surprisingly, BPTT can learn about {\it for
  loops} and can also learn to perform addition and multiplication of
strings containing several decimal digits. Achieving this kind of behaviour
seemed ludicrously optimistic when BPTT was first proposed.

@@

Human vision is an active process in which intelligent fixations are used
to ensure that only a tiny fraction of the optic array is ever processed at
high resolution. We expect much of the future progress in vision,
especially for video understanding or robotics, to come from systems that
are trained end-to-end and combine deep convolutional networks with
recurrent nets that use reinforcement learning to decide where to
look\cite{}.  Such systems already outperform passive vision
systems\cite{ba\&mnih}.  Similarly, when understanding sentences or whole
documents we expect future systems to learn strategies for selectively
attending to one part at a time~\cite{Bahdanau-et-al-arxiv2014}. We also expect
that in the next few years deep recurrent neural networks will become far
better at understanding the meaning of documents, including the way in
which the thoughts expressed by the sentences are related to one another.

This review has focused on supervised learning because of its recent
achievements and we have ignored much interesting work on unsupervised
learning procedures for deep neural networks~\cite{deep-boltz, wake-sleep,
  google-cat, denoising-auto, ranzato-pami, more recent}. In the long run,
however, we believe unsupervised learning will be essential for allowing
deep networks to generalize well by creating internal representations that
untangle the many different factors that interact to produce images, speech
and documents~\cite{Yann, Yoshua: untangling papers,Bengio-Courville-Vincent-TPAMI2013}.

Finally, we believe that the recent successes of deep learning using
stochastic gradient descent are only the beginning.  As pointed out
in~\citet{Bengio+Lecun-chapter2007-small}, end-to-end learning
systems that have many parameters and few built-in assumptions can make use
of more data and more compute power far more easily than systems that
require hand-engineering of domain-specific knowledge.
